Efficient Computation of Extended Sources

### 2022-03-28 repro/papers/GEM22/paper.tex
first commit

@@

As is well-known, local optimization methods are the only feasible
approach given the dimensions of a typical instance of \ref{eqn:fwi},
and those have a tendency to stall due to ``cycle-skipping''. See for
example \cite{VirieuxOperto:09} and many references cited there. Source
extension is one approach to avoiding this problem. It consists in
imposing the wave equation as a soft as opposed to hard constraint, by
allowing the source field $\bff$ to have more degrees of freedom than
is permitted by a faithful model model of the seismic experiment, and
constraining these additional degrees of freedom by means of an
additive penalty modifying the probem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

@@

The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
subproblem is a linear least squares problem. Under some additional
assumptions to be described below, I shall show how to construct an
accurate approximate solution operator for problem
\ref{eqn:esis}. This approximate solution operator may be used to
accelerate Krylov space methods for the solution of the source
subproblem \ref{eqn:esis}. Numerical examples suggest the
effectiveness of this acceleration.

I will fully describe a preconditioner for a special
case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
acoustic field, $L[{\bf c}]$ is the wave operator of linear
acoustodynamics, the spatial positions of traces extracted by $P$ lie
on a plane, and the positions at which the extended
source $\bff$ is nonzero lie on another, parallel, plane. The data and
sources are further limited to pressure traces and constitutive law
defects (``pressure sources''). While this
``crosswell'' configuration simplifies the analysis underlying the
construction of approximate solutions for the source subproblem
\ref{eqn:esis}, it is only one of many transmission configurations for
which similar developments are possible. Perhaps the most important
alternative example is the diving wave configuration, which plays a
central role in contemporary FWI. I will discuss the generalization of
surface source extention to diving wave inversion at the end of the
paper.

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. These are essentially the same as the ``hyperbolic
Dirichlet-to-Neumann'' map that plays a prominent role in
thermoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

The obvious computation of the pressure-to-source map
- prescribe the pressure, solve the wave equation with this boundary
condition, read off the equivalent source - suffers from intrinsic
numerical inaccuracy. I suggest an alternative computationally
feasible approach.

@@

A technical difficulty with spatial sampling must be
addressed. Acoustic field energy is defined in terms of the mean
squares of the pressure and particle velocity fields. However fields
with finite mean square do not in general have well-defined
restrictions to lower-dimensional sets: for example, the pressure
field of a finite-energy acoustic vector field may not have a
well-defined restriction to the space-time plane $z=z_r$. This
phenomenon is related to the ill-posedness of wave equations as
evolution equations in spatial variables, an observation attributed to
Hadamard (see \cite{CourHil:62}, Chapter 6, section 17, also
\cite{Payn:75,Symes:83}). Some constraint on the acoustic field,
beyond finite energy, is mandatory in formulating the inverse problems
\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
the ``crosswell'' geometry of this paper is that high-frequency energy
{\em not} travel along rays parallel or nearly parallel to the
surfaces $z=z_s, z=z_r$. I will call fields with this property {\em
  downgoing} (even though the concept also encompasses {\em upcoming}
fields). \cite{BaoSy:91b} give a mathematically complete discussion of
downgoing wave field properties.

@@

\section{2D Examples}

\inputdir{project}

To illustrate the structure described in the preceding section, I
introduce two 2D acoustic models, one spatially homogeneous, the other
with highly refractive. The first, homogenous model has $\kappa = 4$
GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
($z$). The second, refractive, model is a perturbation of the first by
a low-velocity acoustic lens positioned in the center of the
rectangle. To produce this structure, the density is chosen
homogeneous as in the first model, while the bulk modulus decreases to
from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.

Discretization is conventional, with a rectangular grid and staggered
finite difference scheme \cite[]{Vir:84} of order 2 in time and
2$k$ in space; for most of the experiments reported below, $k=4$.
Sampling operators such as $P_r$ are implemented via linear
interpolation, and source insertion via adjoint linear interpolation
(as noted above, in the continuum limit, sources are represented via
adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
$\Delta x = 20$ m, and limit the temporal frequency of the computed
traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
maintaining 8 samples per wavelength. All computations are carried out
in single precision.

\cite{GeoPros:11} gives a description of the code
implementation, out-of-date in a few respects but overall accurate.

The horizontal line of receivers sits at depth $z_r = $ 1000 m, as
shown in Figure \ref{fig:bml0}. Receiver $x$ ranges from $2000$ to
$6000$ m. A single point (physical) source appears in these
experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
in the figure. The source time function is a bandpass filter with
corner frequencies $1, 2.5, 7.5, 12.5$ Hz for the lowest frequency
source, and scaled as appropriate for examples with higher frequency
and finer sampling.  Extended sources are confined to the horizontal
line through the physical source position, that is $z_s = 3000$ m,
over a 4 km interval starting at $x_r=$ 2000 m. Note that we have
reversed the order relation between $z_s$ and $z_r$ described in the
text ($z_s<z_r$). This difference is immaterial.

The point source pressure data generated by this configuration for the
homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.

While inversion of pressure data is the main object of this exercise,
normal velocity ($v_z$) data will play an important role, so I display
the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.

\section{Adjoints}

@@

The implemented code uses the discrete adjoint state method and auto-generated code \cite[]{TapenadeRef13}, to
assure that the computed adjoint operators are adjoint at the level of
machine precision to the computed operators. The reverse-time storage
issue is resolved through the optimal checkpointing technique
\cite[]{Griewank:book,Symes:06a-pub}, again without loss of precision.

Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
\begin{itemize}
\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
\end{itemize}
Ideally, the differences of these inner products should be at most a
relatively small multiple of machine precision, {\em relative} to the
products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
computationally inaccessible, so instead I used the smaller quantities
$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
overestimating the relative error between the inner products. In all
cases, over a very large number of random inputs, the largest observed
relative error estimate was $O(10^{-9})$, well under the appropriate limit for
single precision.

@@

Since the system \ref{eqn:awepm} has a unique solution by standard
theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
prescribe both pressure and normal velocity on a surface such as
$z=z_s$. So the columns of the matrix operator
${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
section I will explain this relation; it involves the {\em
  pressure-to-source} map. This operator also turns out to be the
principle component of a preconditioning strategy for iterative
solution of the optimization problem \ref{eqn:esis}, so I will devote
some effort to its proper definition. It is closely related to the
Dirichlet-to-Neumann operator mentioned in the introduction.

@@

To give an example of \ref{eqn:snull} in action, it is necessary to
create downgoing data and the action of the operator $\Lambda^{\pm}$
on a downdoing field data $f_s$ on $z=z_s$. Note that a point source
on $z=z_s$ creates high frequency energy traveling on rays parallel
and nearly parallel to $z=z_s$, so that won't do. One simple approach
is to create pressure and normal velocity data on $z=z_s$ by placing a
point source at a depth $z_d<z_s$. Since the examples used here are
homogeneous in $z<z_s$, the traces extracted from the resulting
causal pressure and velocity data on $z=z_s$ are {\em a priori}
downgoing, hence related by the operator $\Lambda^+_{z_s}$.

Since the
mechanical parameters in the homogeneous and lens models are the same
for $z<z_s$, and no rays return to this zone in either model, these
data are asymptotically the same for both models, and I show only the
homogenous medium results. Regard these gathers as the trace
$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
Equations \ref{eqn:tracejump10} and
\ref{eqn:tracejump20} show that these differ by a factor of -2 from
source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.

These source functions satisfy
the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in $z>z_s$. Figures
\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} (lens model) show the pressure
gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
for the two choices of extended source at $z=z_s$, on the same color
scale. The obvious similarity between the fields generated by the two
extended sources,
predicted by equation \ref{eqn:snull}, is confirmed by trace
comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltphh0tr81}
and
\ref{fig:drecplh0tr21}-\ref{fig:daltplh0tr81}.

@@

I have applied this procedure to the pressure gathers shown in Figures
\ref{fig:drecphh0} and \ref{fig:drecplh0}, generated by a point source
at $z_d=3500$ m, $x_s=3500$ m, propagating in the homogeneous and lens
models respectively. The source time dependence is again a [1, 2.5,
10, 12.5] Hz trapezoidal bandpass filter. As noted earlier, in the
region $z < z_s=3000 m$, these fields are downgoing and can be
regarded as the result of either pressure or velocity source at $z=z_s$: the
pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
pressure gather depicted in Figure \ref{fig:dsrcphh0}.

Figure \ref{fig:dinvhshh0} shows the approximate inversion (via the
first equation in display \ref{eqn:approxinv}) of the
pressure gather shown in Figure \ref{fig:drecphh0}, inverted in the homogeneous model
used to generate the data. The result is a scaled, dip-filtered
version of the pressure source shown in Figure \ref{fig:dhshh0}.
The dip filter effect is unavoidable and is caused by the aperture
limitation of the acquisition geometry: the steeper dips in the source
gather (Figure \ref{fig:dhshh0}) do not contribute to the data, and are not
recovered in the inversion.

To see that this is indeed an inversion, I apply $S^{+}_{z_s,z_r}$ to
the source shown in Figure \ref{fig:dinvhshh0}, to produce the
re-simulation shown in Figure \ref{fig:drerecphh0}. Simulation is carried
out in the homogeneous model used in the inversion. Comparison with the
gather in Figure \ref{fig:drecphh0} via a difference plot
\ref{fig:ddiffrecphh0} verifies that a very close match has been
obtained, so the source in Figure \ref{fig:dinvhshh0} is indeed a
satisfactory inversion, even it differs considerably from the original
point source: the difference lies in an approximate null space of the modeling
operator, arising from the limited aperture of the acquisition
geometry.

The remainder of this section is devoted to illustrating two important
features of the approximate inversion developed above. The
first is the validity of the inversion approach given in display
\ref{eqn:approxinv} so long as the fields generated by the
model used in simulation and inversion are downgoing. The models
may be quite far from homogeneous, and even generate triplications,
whereas the inversions remain accurate modulo the dip filtering effect
of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
inverted from the data Figure
\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
this data is generated by propagating a point source at $z_d=3500$ m
in the lens model. The inversion uses the same model. Figure
\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
source at $z=z_s=3000$ m
(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
the same way that the sources in Figures \ref{fig:dhshh0} and
\ref{fig:dinvhshh0} are related. The inverted source is an accurate
inversion: it generates the data depicted in Figure
\ref{fig:drerecpll0}, which is very close to the input data (Figure
\ref{fig:drecplh0}). The difference is displayed on the same color
scale in Figure \ref{fig:ddiffrecpll0}.

The second feature is a decoupling of the models used (implicitly or
otherwise) in simulation and those used in inversion: it is possible
to {\em accurately invert pressure data for pressure source using the
  wrong model}, in the sense that the inverted source will
generate an accurate recovery of the input data re-simulated in the
same (wrong) model. For example, inversion of the data of the previous
example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
that is very far from a dip-filtered version of the source used
generate the data (Figure \ref{fig:dhshh0}). However, this result is an
accurate inversion: re-simulation {\em using the same (homogeneous)
  model as used in the inversion} results in accurate recovery of the
input pressure gather (Figure \ref{fig:drecplh0}). The difference is
shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.

@@

A minor extension of an earlier example, based on the gathers
displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
\ref{eqn:sv}.
The pressure gathers at $z=z_r$ produced by application of
$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
and \ref{fig:daltphh0}), and are identical to several digits. The same 
is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
expresses the interchangeability of pressure and velocity sources,
related by $\Lambda$.

Relation \ref{eqn:tracejump10} characterizes
$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
of downgoing fields restricted to $z=z_s$. That is, the pressure
source gather
$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
pressure gather in Figure \ref{fig:dsrcphh0} under
$\Lambda^{+}_{z_s}$.
The pressure gather
\ref{fig:dfwdphh0} is the image of this pressure source gather under
$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
$-2$ produces
$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
hand, figure \ref{fig:daltvzhh0} shows the result of applying
$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
are essentially identical, the relation \ref{eqn:sv} holds for this
example.

@@

To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
operator
\[
  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
  \Lambda^+_{z_r})
\]
to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecphh0}), in which
$h_s$ is the dip-filtered version (Figure \ref{fig:dinvhshh0}) of the
downgoing source created earlier (Figure \ref{fig:dhshh0}), produced
by inversion of the data in Figure \ref{fig:drecphh0} via the first
equation in display \ref{eqn:approxinv}, propagation in the homogenous
model.  The operator above is computed via the technique explained in
the section ``Computing and Symmetrizing $\Lambda$'', below, using
auxiliary receiver arrays 100 m above the data source and receiver arrays.

The output is shown in Figure
\ref{fig:lamsstlamrdrecphh0}. The difference with the actual source is
shown in Figure \ref{fig:difflamsstlamrdrecphh0}. A similar exercise
using the lens data \ref{fig:drecplh0} but inversion in the homogenous
model produces the extended source depicted in
\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
via equation \ref{eqn:approxinv} by the gather displayed in Figure
\ref{fig:difflamsstlamrdrecplh0}.

@@

[Example]

@@

Figure \ref{fig:preddinvhshh0} shows the pressure source gather
produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
application of the operator on the right-hand side of formula
\ref{eqn:lamident} to the source surface pressure gather depicted in
\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
implicit in the prescription \ref{eqn:lamident}. The scale is the same
as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
  \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
  similarity is quantitative. The recovered source gather (Figure
  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
  difference with the point source gather (Figure \ref{fig:drecphh0})
  plotted on the same color scale in Figure
  \ref{fig:ddiffpredhsrecphh0}.

Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
the same comparison for the $\Lambda^+_{z_s}$ approximation in
equation \ref{eqn:lamident}, with all propagation taking place in the
lens model.

@@

To illustrate these points, I have re-computed the source gather
estimates shown in Figure \ref{fig:preddinvhshh0}. In place of the receiver geometry in the
target data (Figure \ref{fig:drecphh0}), I
place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
source surface at $z_s=3000$ m. For the discretization used to create
the examples shown so far, that is just a 5 gridpoint difference in
depth, as opposed to 100 gridpoints between the source and receiver
depths for examples such as shown in Figure \ref{fig:recphh0} or
\ref{fig:drecphh0}.

Note that the homogeneous and lens models are identical in the depth
range $2900 < z < 3000$ m, so the computed $\Lambda^+_{z_s}$ operators
will be the same for both models. Hence I show only results for the
homogenous model.

The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamident}
for this configuration is evaluated in the same way as before in
Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}. The effect of aperture
limitation is clearly diminished: the second figure in this series
compares the full-aperture pressure source gather (Figure
\ref{fig:dhshh0}) with the 
image of the corresponding pressure gather (Figure \ref{fig:dsrcphh0})
under the approximation to $\Lambda_{z_s}^+$, and the last figures
show that the approximated source gathers accurately predict the
point-source pressure gather at the receiver datum $z_r=1000$ m.

@@

>>>>

\inputdir{../sseprecond/project}

@@

As is well-known, local optimization methods are the only feasible
approach given the dimensions of a typical instance of \ref{eqn:fwi},
and those have a tendency to stall due to ``cycle-skipping''. See for
example \cite{VirieuxOperto:09} and many references cited there. Source
extension is one approach to avoiding this problem. It consists in
imposing the wave equation as a soft as opposed to hard constraint, by
allowing the source field $\bff$ to have more degrees of freedom than
is permitted by a faithful model model of the seismic experiment, and
constraining these additional degrees of freedom by means of an
additive quadratic penalty modifying the problem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

@@

The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
subproblem is a linear least squares problem. Under some additional
assumptions to be described below, I shall show how to construct an
accurate approximate solution operator for problem
\ref{eqn:esis}. This approximate solution operator may be used to
accelerate Krylov space methods for the solution of the surface source
subproblem \ref{eqn:esis}. Numerical examples suggest the
effectiveness of this acceleration.

I will fully describe a preconditioner for a special
case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
acoustic field, $L[{\bf c}]$ is the wave operator of linear
acoustodynamics, the spatial positions of traces extracted by $P$ lie
on a depth plane $z=z_r$, and the positions at which the extended
source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
%That is, sources take the form of a combination of constitutive law
%defect and normal force,
%\begin{equation}
%  \label{eqn:surfsrc}
%  \bff(x,y,z,t) = h_s(x,y,t)\delta(z-z_s)(1,0,0,0)^T + f_s(x,y,z,t)(0,0,0,1)^T
%\end{equation}
%and the sampling operator $P$ is defined by
%\begin{equation}
%  \label{eqn:surfsam}
%  P\bu(x,y,t) =
%\end{equation}
%supported on $z=z_s$, and similar
%surface vertical loads.
This
``crosswell'' configuration simplifies the analysis underlying the
construction of approximate solutions for the source subproblem
\ref{eqn:esis}. It is only one of many transmission configurations for
which similar developments are possible. Perhaps the most important
alternative example is the diving wave configuration, which plays a
central role in contemporary FWI.

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. This pressure-to-source map is closely related to the ``hyperbolic
Dirichlet-to-Neumann'' operator that plays a prominent role in
photoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

The obvious computation of the pressure-to-source map - prescribe the
pressure, solve the wave equation with this boundary condition, read
off the equivalent source - suffers from intrinsic numerical
inaccuracy. I suggest an alternative computationally feasible
approach, via economical short-distance wave propagation. Since the
map is symmetric only in an approximate, asymptotic sense, it must be
symmetrized for use as a Krylov preconditioner. I describe a
symmetrization procedure that requires no further wave computations
beyond those necessary to compute the action of the operator itself.

The discussion in this paper is formal and incomplete, in the sense
that some important mathematical underpinnings are only
sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
it mapped square integrable surface sources to square integrable sampled
data. This is not true in full generality: while the surface source
problem has distribution solutions, they are not generally square
integrable (finite acoustic field energy). Even if the solutions have
finite energy, they do not in general have well-defined
restrictions to lower-dimensional sets. In other words, the action of
the sampling
operator $P$ on the space-time plane $z=z_r$ is not well-defined for arbitrary
finite-energy acoustic fields. Thus the modeling operator envisioned above
may not be well-defined.

This
phenomenon is related to the ill-posedness of wave equations as
evolution equations in spatial variables, an observation attributed to
Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). Some constraint on the acoustic field,
beyond finite energy, is mandatory in any precise mathematical formulation the inverse problems
\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
the ``crosswell'' geometry of this paper is that high-frequency energy
travel {\em only} along rays crossing the
surfaces $z=z_s, z=z_r$ transversally. That is, source functions on
$z=z_s$ generate waves with energy traveling along rays leaving the surface at a non-zero angle, and
energy arrives at the recording surface $z=z_r$ along rays making
non-zero angle with it. I will call sources, sampled data, and
acoustic fields with this property {\em
  downgoing} (even though the concept also encompasses {\em upcoming}
propagation). Note that the downgoing property restricts the behaviour of
acoustic fields near the source and receiver surfaces - what the
fields do elsewhere is their own business.

Several works have explored the mathematics of the
downgoing condition and its consequences in the context of the scalar second order wave equation, see for instance
\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}. 
Since the purpose of this paper is to explore the algorithmic
consequences of the mathematical structure implied by the downgoing
condition, I leave the elaboration of the mathematical details to
another place.

The next section defines the modeling operator $PL[{\bf c}]$, its
adjoint, and important specializations (pressure vs. normal velocity
sources and data). The sections to follow define the
source-to-pressure operator, construct an approximate inverse of the
modeling operator by time reversal (as suggested by work in
photoacoustic tomography), use the source-to-pressure operator to
express the approximate inverse as the modeling operator adjoint in
weighted norms (thus establishing that the modeling operator is {\em
  unitary} in the sense of these norms), explain how to use this
construction to precondition Conjugate Gradient iteration, and
organize the preconditioning computation so as to involve only one
wave propagation calculation. The penultimate section displays simple
2D numerical examples of all of the key steps, culminating in a
comparison of straight vs. preconditioned CG iteration. The paper ends
with a brief discussion-and-conclusion section, reviewing what has
been accomplished and listing a few of the many questions left open.

@@

@@

@@

@@

Since the system \ref{eqn:awepm} has a unique solution by standard
theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
prescribe both pressure and normal velocity on a surface such as
$z=z_s$. So the columns of the matrix operator
${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
section I will explain this relation; it involves the {\em
  pressure-to-source} map. This operator also turns out to be the
principal component of a preconditioning strategy for iterative
solution of the optimization problem \ref{eqn:esis}, so I will devote
some effort to its proper definition. It is closely related to the
Dirichlet-to-Neumann operator mentioned in the introduction.

@@

@@

@@

@@

@@

@@

@@

@@

\section{Numerical Examples}
This section illustrates the most important conclusions developed in
the preceding sections by finite difference wavefield simulation.

\subsection{Synthetic models and simulation}
To illustrate the structure described in the preceding section, I
introduce two 2D acoustic models, one spatially homogeneous, the other
highly refractive. The first, homogenous model has $\kappa = 4$
GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
($z$). The second, refractive, model is a perturbation of the first by
a low-velocity acoustic lens positioned in the center of the
rectangle. To produce this structure, the density is chosen
homogeneous as in the first model, while the bulk modulus decreases to
from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.

Discretization is conventional, with a rectangular grid and staggered
finite difference scheme \cite[]{Vir:84} of order 2 in time and
2$k$ in space; for most of the experiments reported below,
$k=4$. Absorbing boudary conditions of perfectly matched layer type
are applied at all boundaries of the simulation rectangle \cite[]{Habashy:07}.
Sampling operators such as $P_r$ are implemented via linear
interpolation, and source insertion via adjoint linear interpolation
(as noted above, in the continuum limit, sources are represented via
adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
$\Delta x = 20$ m, and limit the temporal frequency of the computed
traces to 12.5 Hz.

\cite{GeoPros:11} gives a description of the code
implementation, out-of-date in a few respects but overall accurate.
The implementation uses the discrete adjoint state method and
auto-generated code \cite[]{TapenadeRef13}, to assure that the
computed adjoint operators are adjoint at the level of machine
precision to the computed operators. The reverse-time storage issue is
resolved through the optimal checkpointing technique
\cite[]{Griewank:book,Symes:06a-pub}, again without loss of
precision. This procedure results in computed adjoints for
$S^+_{z_s,z_r}$ and other operators that pass usual test for adjoint
accuracy, comparing inner products with pseudorandom input vectors,
with errors well under machine precision.

%Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
%\begin{itemize}
%\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
%  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
%\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
%  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
%\end{itemize}
%Ideally, the differences of these inner products should be at most a
%relatively small multiple of machine precision, {\em relative} to the
%products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
%$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
%dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
%computationally inaccessible, so instead I used the smaller quantities
%$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
%overestimating the relative error between the inner products. In all
%cases, over a very large number of random inputs, the largest observed
%relative error estimate was $O(10^{-9})$, well under the appropriate limit for
%single precision.

%We also explore the dependence of a few results on frequency, using
%$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
%maintaining 8 samples per wavelength. All computations are carried out
%in single precision.

The horizontal line of receivers sits at depth $z_r = $ 1000 m, the
(extended) sources at $z_s=3000$ m.  %as
% shown in Figure \ref{fig:bml0}.
Source and receiver $x$ ranges from $2000$ to
$6000$ m. Note that we have
reversed the order relation between $z_s$ and $z_r$ described in the
text ($z_s<z_r$). This difference is immaterial for the purpose of
illustrating the mathematical structures developed in the preceding paragraphs.
%A single point (physical) source appears in these
%experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
%in the figure.

%Extended sources are confined to the horizontal
%line through the physical source position, that is $z_s = 3000$ m,
%over a 4 km interval starting at $x_r=$ 2000 m..

%The point source pressure data generated by this configuration for the
%homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
%lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.

%While inversion of pressure data is the main object of this exercise,
%normal velocity ($v_z$) data will play an important role, so I display
%the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.

\subsection{Creating downgoing fields}
The downgoing condition constrains high-frequency energy of localized
plane wave components, hence could be enforced by dip
filtering. However, a simpler approach is to construct fields that
must be entirely downgoing at the source and receiver surfaces by
virtue of ray geometry.
